---
title: "pathways Enrichment Analysis on TCGA Breast Cancer"
author: "Parissa Amin"
date: "2023-12-31"
output: pdf_document
---

#  DEA

We first perform DEA as before for ER+ vs ER- subtypes of breast tumors

```{r,  out.width="100%" }
library(fgsea)
library(clusterProfiler)

library(ggplot2)
library(ggpubr)  
library(dplyr)
library(dplyr)
library(DESeq2)
library(enrichplot)
library(ggupset)
library(tidyverse)
library(org.Hs.eg.db)
library(DOSE)

#############################
exp <- read.delim("D:/BulkRNA/Data/data_mrna_seq_v2_rsem.txt")
exp<- na.omit(exp)
exp<- data.frame(t(exp))

colnames(exp) <- exp[ 1,]
exp<- exp[-1:-2, ]

################PatientInfo 
PatientInfo <- read.delim("D:/BulkRNA/Data/tcga_clinical_data.tsv")
PatientInfo$Sample.ID <- gsub("-" ,"." , PatientInfo$Sample.ID)

## Defining a new column named Er.status
exp$ER.status <- PatientInfo$ER.Status.By.IHC[match(rownames(exp), PatientInfo$Sample.ID)]
exp<- na.omit(exp)
exp<- exp[exp$ER.status!="Indeterminate",]


#############Creating a metadata table with information of our samples 
metadata<- as.data.frame(cbind(rownames(exp), exp[,"ER.status"]))
colnames(metadata)<- c("Sample.ID", "ER")

#####Deleting last column(the genes' names) of exp
expression_data<-  exp[,-20531]
expression_data <- data.frame(lapply(expression_data, as.numeric))
expression_data<-as.data.frame(log2(expression_data+1)) # Adding 1 to avoid log(0)
expression_data<- round(expression_data)
expression_data<- as.data.frame(t(expression_data))

######################################Create dds object by DESeqDataSet function

dds<-DESeqDataSetFromMatrix(countData=expression_data,colData=metadata,design=~ER)
dds <- DESeq(dds)
My_Results<- results(dds)
My_Results_df<- as.data.frame(My_Results)
My_Results_df$GeneName <- rownames(My_Results_df)
```


  Selecting DEGs 

```{r}
Genes_in_data <- My_Results_df$GeneName
My_Results_df <- na.omit(My_Results_df)

My_Results_df <- My_Results_df %>% mutate(diffexpressed = case_when(
  log2FoldChange       > 0 & padj < 0.05 ~ 'Up',
  log2FoldChange       < 0 & padj < 0.05 ~ 'Down',
  padj > 0.05 ~ 'NO'))

My_Results_df<-My_Results_df[My_Results_df$diffexpressed !='NO',]
# Split the dataframe into a list of sub-dataframes: upregulated, downregulated genes
deg_results_list <- split(My_Results_df, My_Results_df$diffexpressed)

```

# GSEA

Load MSigDB Hallmark gene sets

```{r}
hallmark_gene_sets <- read.gmt("h.all.v2023.2.Hs.symbols.gmt")
hallmark_gene_sets2 <- hallmark_gene_sets[hallmark_gene_sets$gene %in% Genes_in_data,] 
```

Run clusterProfiler


```{r}
name_of_comparison <- 'ER' 
background_genes <- 'hallmark_gene_sets' # for our filename
bg_genes <- hallmark_gene_sets2
padj_cutoff <- 0.05 # p-adjusted threshold, used to filter out pathways
genecount_cutoff <- 5 # minimum number of genes in the pathway, used to filter out pathways

# Run clusterProfiler on each sub-dataframe
res <- lapply(names(deg_results_list),
              function(x) enricher(gene = deg_results_list[[x]]$GeneName,
                                   TERM2GENE = bg_genes))
names(res) <- names(deg_results_list)
```

  
Convert the list of enrichResults for each sample_pattern to a dataframe with the pathways 

```{r}
res_df <- lapply(names(res), function(x) rbind(res[[x]]@result))
names(res_df) <- names(res)
res_df <- do.call(rbind, res_df)
head(res_df)

 
res_df <- res_df %>% mutate(minuslog10padj = -log10(p.adjust) )
```


Subset to those pathways that have p adj < cutoff and gene count > cutoff (you can also do this in the enricher function)
```{r}
target_pws <- unique(res_df$ID[res_df$p.adjust < padj_cutoff & res_df$Count > genecount_cutoff]) # select only target pathways have p adjusted < 0.05 and at least 6 genes
res_df <- res_df[res_df$ID %in% target_pws, ]


```

 
# PLOT

enrichres object 

```{r}
enrichres <- new("enrichResult",
                 readable = FALSE,
                 result = res_df,
                 pvalueCutoff = 0.05,
                 pAdjustMethod = "BH",
                 qvalueCutoff = 0.2,
                 organism = "human",
                 ontology = "UNKNOWN",
                 gene =Genes_in_data,
                 keytype = "UNKNOWN",
                 universe = unique(bg_genes$gene),
                 gene2Symbol = character(0),
                 geneSets = bg_genes)
class(enrichres) 
```
 
 
 Barplot 

```{r}

barplot(enrichres, showCategory = 20) 
```

Sorted Barplot 

```{r}

mutate(enrichres, qscore = -log(p.adjust, base = 10)) %>% 
  barplot(x = "qscore")
```

dot plot

```{r}


dotplot(enrichres, showCategory = 15) + ggtitle("ER+ vs ER-") 
```
 


# Conclusion
We ran GSEA to compare ER+ and ER- tumors in breast cancer. As expected, alongside many other pathways,  Estrogen Receptor pathways were significantly enriched in ER+ tumors. 



